Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-09-26 demystify.tex
updates

\subsection{Self-written Letters are Bad} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot  (because the student often has little experience in writing LoRs) and  \red{red flag} such letters.

Most well-known and well-respected profs. would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). This might be a common practice at their university or the students do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer, relatively unknown in the field, or simply do not know/care enough about you.  In any of these cases, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

@@

Many international students try to get letters from high-ranking admin people in their universities, e.g., department head, dean, or director.  These people, while well-known and respected in your university, might not be too active in research (e.g., they focus on administration) and thus are not recognized by adcom members, who often recognize good researchers (even if they are not in their field'', but not admin people. So it's fine to to have `less well-known'' people write for you as long as they can write a good letter about your research ability.

\begin{commentbox}[Vu:]
  In my experience in reviewing applications, letters from admin people are often generic and do not provide much value.  If you are in this situation, you should find someone else to write for you.
\end{commentbox}

@@

\subsection{Help your LOR Writer} As mentioned above, do not write your own letter and generic letters do not give much value. One thing you can do to help your writer to write a strong, customized LoR for you is to provide details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
Sometimes your writer will explicitly ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

@@

While you might not have control over LORs (\autoref{sec:lor}) or where your go to school (\autoref{sec:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference.

In your statement, you have the opportunity to connect directly with adcom members and make your application stand out and unique. Your SOP can convince adcom members that you would \emph{fit} the program, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

In your SOP, focus on research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved certain performance improvement over the state of the art. Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

Finally, you should get your SOP reviewed by your LoR writers, professors (especially those who have served in adcom), or even postdocs or PhD students (because they have been through this process).

@@

A common question is whether you should be concern and explain bad grades in relevant courses in your SOP.  If you have just a few, then they do not matter much, so don't spend much time explaining them (many adcom reviewers themselves have bad grades in relevent courses!). But if you have many bad grades for an entire semester or year due to some specific reasons, then you may be you should explain them in your SOP.

@@

@@

After you apply, you might get interviews. The most common case is that a prof. is interested in working with you and wants to chat with you, e.g., to offer RA (\autoref{sec:ra}). In some cases, the interview is done by several professors, e.g., to see if a student fits in their group or to recruit a very strong student to their program. The interview can also be done by adcom members to see if the student fits their program.

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. 
You might be asked to talk about your research experience and interests and to read a paper and discuss it. In some rare cases you might also be asked to solve a problem, e.g., a coding problem (one of my colleagues at Mason often does this).
The interview is also a good opportunity for you to ask questions, e.g., about the prof, their group, the CS PhD program, and the university.

@@

  However, be aware that heathcare services can still be expensive even with insurance. So  know what your insurance covers and be prepared for unexpected costs.  Moreover, healthcare system has many confusing jargons such as HMO, PPO, deductibles, copays, and coinsurance (try take a look at the Explanation of Benefits or EoB statement you received from your insurance company). It's arguably the most complicated system in the US and even Americans often do not understand it (and politicians often exploit this to their advantage).  So do not hesitate to ask your HR or the insurance company for help.

>>>>

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
 Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot  (because the student often has little experience in writing LoRs) and  \red{red flag} such letters.

Most well-known and well-respected profs would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). This might be a common practice at their university or the students do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer, relatively unknown in the field, or simply do not know/care enough about you.  In any of these cases, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

@@

Many international students try to get letters from high-ranking admin people in their universities such as department head, dean, or director. The students often mistakenly believe that these letters are more valuable because such people are important in their universities.  However, these people, while well-known and respected in your university, might not be active in research (e.g., they focus on administration) and thus are not recognized by adcom members. In fact, to adcom members, such a letter has little value if the writer does not know you well enough (e.g., did not work with you) to talk about your research ability.

\begin{commentbox}[Vu:]
  In my experience in reviewing applications, letters from admin people are often generic and do not provide much value. 
  In many case, the letter reads like it was written by a student (\autoref{sec:self-letters}), and thus is a red flag.  So if you are in this situation, you should find someone else to write for you.
    So if you are in this situation, find someone else to write for you.
\end{commentbox}

@@

\subsection{Help your LOR Writer} As mentioned above, do not write your own letter and generic letters do not give much value. However, you can help your writer to write a strong, customized LoR for you by providing details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

Sometimes your writer will explicitly ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

@@

While you might not have control over LORs (\autoref{sec:lor}) or where your go to school (\autoref{sec:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference. A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).

%In your statement, you have the opportunity to connect directly with adcom members and make your application stand out and unique. Your SOP can convince adcom members that you would \emph{fit} their program, even if you don't have very strong research experience.
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

In your SOP, focus on research potential and convince us through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved certain performance improvement over the state of the art. Also, do not use more than two pages (references can be on the third page). Many SOP samples are \href{https://cs-sop.org/}{available here}.

@@

Finally, you should get your SOP reviewed by your LoR writers and professors, especially those who have served in adcom, or even postdocs or PhD students as they have been through this process.

@@

A common question is whether you should be concern and explain bad grades in relevant courses in your SOP.  If you have just a few, then they do not matter much, so don't spend much time explaining them (many adcom reviewers themselves have bad grades in relevent courses!). But if you have many bad grades for an entire semester or year due to some specific reasons, then you should explain them in your SOP.

@@

\begin{commentbox}[Vu:]
    Unlike a job application, CV is not as important for PhD admission. We care more about your LOR, SOP, etc.  So do not spend too much time on your CV, just make sure it is easy to skim through and well-organized around research activities and achievements.
\end{commentbox}

@@

After you apply, you might get interviews. The most common case is that a prof. is interested in working with you and wants to chat with you, e.g., to offer RA (\autoref{sec:ra}). In some cases, the interview is done by several professors, e.g., to see if a student fits in their group or to recruit a very strong student to their program.

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. 
You might be asked to talk about your research experience and interests and to read a paper and discuss it. In some rare cases you might also be asked to solve a problem (one of my colleagues at Mason often does coding interview).
The interview is also a good opportunity for you to ask questions, e.g., about the prof, their group, the CS PhD program, and the university.

@@

  However, be aware that heathcare services can still be expensive even with insurance. So  know what your insurance covers and be prepared for unexpected costs.  Moreover, healthcare system has many confusing jargons such as HMO, PPO, deductibles, co-pays, and coinsurance (try take a look at the Explanation of Benefits or EoB statement you received from your insurance company). It's arguably the most complicated system in the US and even Americans often do not understand it (and politicians often exploit this to their advantage).  So do not hesitate to ask your HR or the insurance company for help.
